---
title: "AndrewBowen_Data606_LAb1"
author: "Andrew Bowen"
date: "2022-08-26"
output:
  pdf_document: default
  html_document: default
---

```{r, echo=FALSE}
library(openintro)
library(ggplot2)
library(dplyr) # used for piping operator later
```

```{r}
data('arbuthnot', package='openintro')

head(arbuthnot, 10)
```
**Exercise 1**
Just want to see the count of girls baptized
```{r}
arbuthnot$girls
```

## Plotting our DF
```{r}
ggplot(data=arbuthnot, aes(x=year, y=boys)) + geom_point()
```
Same plot as above but in line format


**Exercise 2**
Plotting a line graph the # of girls baptized by year below. There is a general upward trend in baptisms. On a smaller timeframe, there was a dip in girl baptisms between 1640 and 1660.
```{r}
ggplot(data=arbuthnot, aes(x=year, y=girls)) + geom_line()
```

```{r}
arbuthnot$boys + arbuthnot$girls
```

Adding `total` field to dataframe
```{r}
arbuthnot <- arbuthnot %>% mutate(total = boys + girls)

head(arbuthnot, 10)
```

Plotting `total` over time
```{r}
ggplot(data=arbuthnot, aes(x=year, y=total)) + geom_line()
```
Calculationg boy-to-girl ratio
```{r}
arbuthnot <- arbuthnot %>% mutate(boy_to_girl_ratio = boys / girls)
```

**Exercise 3**
Calculating the ratio of boys baptized to total.
```{r}
arbuthnot <- arbuthnot %>% mutate(boy_ratio = boys / total)
```

Finding years where more boys were present than girls. Adding in a boolean flag column to represent
```{r}
arbuthnot <- arbuthnot %>% mutate(more_boys = boys > girls)
```

Let's take a look at our DF with all our added columns!
```{r}
head(arbuthnot, 10)
```

---
# More Practice
```{r}
data('present', package='openintro')
```


```{r}
head(present, 10)
```

**Exercise 4**

Let's see our data range in the year column first (using `min` & `max`)
```{r}
print(min(present$year))
print(max(present$year))
```

Finding out our data frame dimensions (nrows x ncols)
```{r}
print(ncol(present))
print(nrow(present))
```

Getting out dataframe columns with the built-in `colnames` function
```{r}
colnames(present)
```


**Exercise 5**
Going to use the median count for boys and girls from each data set (`present` vs `arbuthnot`) to compare magnitudes of counts in each data set
```{r}
p_boys_med = median(present$boys)
p_girls_med = median(present$girls)

a_boys_med = median(arbuthnot$boys)
a_girls_med = median(arbuthnot$girls)

# Calculating ratio of present boy/girl counts
boys_ratio = p_boys_med / a_boys_med
girls_ratio = p_girls_med / a_girls_med
print(boys_ratio)
print(girls_ratio)

```

It looks like the median present day birth counts are ~320 times higher than the birth counts listed in our `arbuthnot` dataset. We used median counts to summarize the data set, so it won't be exactly this ratio for the whole data set, but modern birth counts are significantly higher.

**Exercise 6**
Setting up our boy-girl ratio column
```{r}
present <- present %>% mutate(total = boys + girls)
present <- present %>% mutate(boys_ratio = boys / total)
head(present, 10)
```

Let's plot the ratio of boys born over time in our `present` dataset:
```{r}
ggplot(data=present, aes(x=year, y=boys_ratio)) + geom_line()
```


While the ratio of boys born has stayed over 50%, it is experiencing a downward trend over time since 1940, the beginning of our dataset. The observation of boys being born more than girls from the `arbuthnot` dataset does hold up, but has decreased since 1940 in the US.


**Exercise 7**
```{r}
# Truncating output to 10 rows for readability
head(present %>% arrange(desc(total)), 10)
```

We see the highest number of total births in the US come in 1961 with 4268326 total births (boys & girls). It's interesting to note that 8 of the top 10 years in terms of total births came during the [baby boom](https://en.wikipedia.org/wiki/Baby_boom#United_States) years in the post-war era.

Checking out the help function (`?`) built into R objects.
```{r}
?present
```

