---
title: DATA 606 Data Project Proposal
author: Andrew Bowen
---

### Data Preparation

```{r setup, echo=TRUE, results='hide', warning=FALSE, message=FALSE}
# load data

```


### Research question 

**You should phrase your research question in a way that matches up with the scope of inference your dataset allows for.**

Have NBA players after the "3-point revolution" (2015) really improved at shooting 3-pointers?

### Cases 

**What are the cases, and how many are there?**
We will need to look at the 3-point shooting percentages for every NBA player in every season for which there is data. 


### Data collection 

**Describe the method of data collection.**



### Type of study 

**What type of study is this (observational/experiment)?**



### Data Source 

**If you collected the data, state self-collected. If not, provide a citation/link.**




### Dependent Variable

**What is the response variable? Is it quantitative or qualitative?**



### Independent Variable(s)




### Relevant summary statistics 

**Provide summary statistics for each the variables. Also include appropriate visualizations related to your research question (e.g. scatter plot, boxplots, etc). This step requires the use of R, hence a code chunk is provided below. Insert more code chunks as needed.**

```{r}

```
