---
title: "DATA 606 Data Project Proposal"
author: "Andrew Bowen"
output: pdf_document
---

### Setup
```{r, message=FALSE}
library(dplyr)
```

### Data Preparation

```{r setup, echo=TRUE, results='hide', warning=FALSE, message=FALSE}
# load data
data_url <- "https://raw.githubusercontent.com/fivethirtyeight/data/master/fifa/fifa_countries_audience.csv"

fifa <- read.csv(data_url)
```


### Research question 

**You should phrase your research question in a way that matches up with the scope of inference your dataset allows for.**
**Research Question**: Is there a stignificant difference in the mean viewership of soccer in Europe vs another continent (confederation)?

### Cases 

**What are the cases, and how many are there?**
There are viewership data for 191 countries. There are 6 confederations in total. We are interested in comparing the European confederation (UEFA) to the other ones.


### Data collection 

**Describe the method of data collection.**
This data was collected from FIFA TV viewership during a world cup. It was cleaned and put into csv format by FiveThirtyEight and lives in their [fifa data GitHub repo](https://github.com/fivethirtyeight/data/tree/master/fifa)


### Type of study 

**What type of study is this (observational/experiment)?**
This is an observational study as the researchers are not actively changing conditions to influence behaviors/outcomes.


### Data Source 

**If you collected the data, state self-collected. If not, provide a citation/link.**
This data comes from a [FiveThirtyeight data set on FIFA viewership](https://github.com/fivethirtyeight/data/blob/master/fifa/fifa_countries_audience.csv). We'll focus on the GDP-weighted TV viewership column `gdp_weighted_share`, as that accounts for population differences.


### Dependent Variable

**What is the response variable? Is it quantitative or qualitative?**
The dependent/response variable is the `gdp_weighted_share`, which is a country's GDP-weighted audience share (as a percentage) of all viewers of a world cup. This is qualitative data


### Independent Variable(s)
Our independent variable is the Confederation (UEFA - Europe, CONCACAF - North America, etc.) to which a country belongs. This is categorical.


### Relevant summary statistics 

**Provide summary statistics for each the variables. Also include appropriate visualizations related to your research question (e.g. scatter plot, boxplots, etc). This step requires the use of R, hence a code chunk is provided below. Insert more code chunks as needed.**
```{r}
library(ggplot2)

ggplot(fifa, aes(x=gdp_weighted_share, group=confederation, color=confederation)) + geom_histogram()

```
These distributions do not appear normal, so we'll use the median as our summary stat.

```{r}
fifa %>%
    group_by(confederation) %>%
    summarise(median_gdp_weighted_share = median(gdp_weighted_share))

```
### Hypothesis Testing

- $H_0$: The mean GDP-weighted viewership share of the world cup in Europe is *not higher*  than that of other confederations
- $H_a$: The mean GDP-weighter viewership share of the world cup is *higher* than that of other confederations


Running a t-test between these two groups with a significance level $\alpha = 0.05$
```{r}
# Filtering into our two groups: Europe vs not Europe.
europe <- fifa %>% filter(confederation == "UEFA")
other_countries <- fifa %>% filter(confederation != "UEFA")

# Running one-tailed t-test using R built-in
t.test(europe$gdp_weighted_share, other_countries$gdp_weighted_share, alternative="greater")
```

Since our p-value is less than our significance level ($\alpha = 0.05$), we can reject the null hypothesis and state that the average weighted viewership of the world cup is higher in Europe than in other confederations.