---
title: "DATA 606 Final Presentation Slides"
author: "Andrew Bowen"
date: "2022-11-29"
output: ioslides_presentation
---


## Research Question
**Is there a significant difference in the mean viewership of soccer in Europe vs another continent (confederation)?**


```{r, echo=FALSE, message=FALSE}
library(ggplot2)
library(dplyr)
library(maps)
library(stringr)
```


## Data

- Viewership data for the 2010 World Cup
- Collected into a [csv file by FiveThirtyEight](https://github.com/fivethirtyeight/data/tree/master/fifa)
- 5 fields: 
  - `country`
  - `confederation`
  - `tv_audience_share`
  - `population_share`
  - `gdp_weighted_share`

Read in the csv into a dataframe
```{r setup, echo=TRUE, results='hide', warning=FALSE, message=FALSE}
# load data
data_url <- "https://raw.githubusercontent.com/fivethirtyeight/data/master/fifa/fifa_countries_audience.csv"

fifa <- read.csv(data_url)
```

```{r, echo=FALSE}
# Adding continent column for cleaner plots/readability
confederation <- c("CONCACAF", "UEFA", "CONMEBOL", "AFC", "CAF", "OFC")
continent <- c("North America", "Europe", "South America", "Asia", "Africa", "Oceania")
continent_map <- data.frame(confederation, continent)

fifa <- merge(fifa, continent_map)

fifa <- fifa %>%
              mutate(european = ifelse(fifa$confederation == "UEFA", TRUE, FALSE))

# Cleaning up country codes
world_map <- map_data("world") %>%
              rename("country" = "region")

# USA named differently in each dataset, replacing to standardize column values (same wqith UK)
world_map$country <- str_replace(world_map$country, "USA", "United States")
world_map$country <- str_replace(world_map$country, "UK", "United Kingdom")

# Joining our world_map geographic data to our World Cup viewership data by country
fifa_coords <- world_map %>% 
  left_join(fifa, by = c("country"))

head(fifa_coords)

```

## Exploratory Data Analysis

```{r, gdp-share-by-confederation, message = FALSE}
# Plotting GDP Weighted share by confederation
ggplot(fifa, aes(x = gdp_weighted_share, fill = confederation)) + 
  geom_histogram(binwidth=1) + 
  facet_grid(confederation ~ .)
```
## TV Audience Share by Confederation

```{r, tv-share-by-confederation, message = FALSE}
ggplot(fifa, aes(x = tv_audience_share, fill = confederation)) + 
  geom_histogram(binwidth=1) + 
  facet_grid(confederation ~ .)
```
## Population Share by Confederation

```{r, pop-share-by-confederation, message = FALSE}
ggplot(fifa, aes(x = population_share, fill = confederation)) + 
  geom_histogram(binwidth=1) + 
  facet_grid(confederation ~ .)
```

## What is a Confederation?

Easier to think of confederations as continents
```{r, confederation-membership}
ggplot(fifa_coords, aes(long, lat, group = group)) +
  geom_polygon(aes( group=group, fill=confederation)) + 
  ggtitle("FIFA Confederation memberdship by country") +
  xlab("Longitude (deg)") + ylab("Latitude (deg)")
```


## TV Audience Share by Country (visualized)

```{r}
# Plotting TV Audience share by country
ggplot(fifa_coords, aes(long, lat, group = group)) +
  geom_polygon(aes( group=group, fill=tv_audience_share)) + 
  ggtitle("TV Audience share (%) of world cup viewership by country") +
  xlab("Longitude (deg)") + ylab("Latitude (deg)")
```

## GDP Weighted SHare by Country (visualized)

```{r, gdp-weighted-share-choropleth}
ggplot(fifa_coords, aes(long, lat, group = group)) +
  geom_polygon(aes( group=group, fill=gdp_weighted_share)) + 
  ggtitle("GDP-weighted share (%) of world cup viewership by country") +
  xlab("Longitude (deg)") + ylab("Latitude (deg)")
```


## Inference
T-test comparing the gdp-adjusted viewership for European vs non-European countries'. We'll be conducting this test with a significance level $\alpha = 0.05$. Our null and alternative hypotheses are listed below:

- $H_0$: The mean GDP-weighted viewership share of the world cup in Europe is *not higher*  than that of other confederations
- $H_a$: The mean GDP-weighter viewership share of the world cup is *higher* than that of other confederations


```{r gdp-weighted-t-test}
europe <- fifa %>% filter(european== TRUE)
other_countries <- fifa %>% filter(european == FALSE)

# Running one-tailed t-test using R built-in
t.test(europe$gdp_weighted_share, other_countries$gdp_weighted_share, alternative="greater")
```


Since our p-value is less than $\alpha = 0.05$, we can reject the null hypothesis and claim that GDP-adjusted average world cup viewership in Europe is higher than that of non-European countries.

## Inference (TV Audience)

Let's also run an t-test on our `tv_audience_share` variable ($\alpha = 0.05$) between European and non-European countries:
- $H_0$: The mean tv audience viewership share of the world cup in Europe is *not higher*  than that of other confederations
- $H_a$: The mean tv audience viewership share of the world cup is *higher* than that of other confederations
```{r tv-audience-t-test}
# Running one-tailed t-test using R built-in method
t.test(europe$tv_audience_share, other_countries$tv_audience_share, alternative="greater")
```


## Inference (ANOVA)
Checking if there is a difference in TV Audience share between confederations
```{r, anova-tv-audience}
confederation_test <- aov(tv_audience_share ~ confederation, data = fifa)

summary(confederation_test)
```

Since our p-value (0.0244) is less than $\alpha = 0.05$, we can reject the null hypothesis and assert that there *is* a statistically significant different in mean TV audience viewership between FIFA confederations. 
