---
title: "DATA 606 Final Project"
author: "Andrew Bowen"
date: "2022-11-22"
output: html_document
bibliography: references.bib
---
### Abstract


### Introduction

Soccer (also called football internationally), is the most popular sport in the world. The FIFA World Cup is the [most-watched sporting event](https://www.nbcsports.com/chicago/world-cup-2022/how-does-world-cups-viewership-tickets-and-attendance-compare-nfls-super) in the world. UEFA, the European confederation for soccer, is generally considered the gold standard of federations. For instance, the [UEFA Champions League](https://www.google.com/search?client=safari&rls=en&q=uefa+champions+league&ie=UTF-8&oe=UTF-8) is considered the highest level of soccer in the world, pitting the best teams in Europe's top leagues against each other. With Europe claiming soccer superiority, does viewership reflect this?

**Research Question**: Is there a significant difference in the mean viewership of soccer in Europe vs another continent (confederation)?

```{r, message=FALSE}
library(ggplot2)
library(dplyr)
library(maps)
library(stringr)
```


### Data

Our data was collected into a csv file by FiveThirtyEight. The original csv file lives in their GitHub repo [here](https://github.com/fivethirtyeight/data/tree/master/fifa). The original file contains 5 fields: 
- `country`
- `confederation`
- `tv_audience_share`
- `population_share`
- `gdp_weighted_share`

We'll read in the csv into a dataframe from the original GitHub repo:
```{r setup, echo=TRUE, results='hide', warning=FALSE, message=FALSE}
# load data
data_url <- "https://raw.githubusercontent.com/fivethirtyeight/data/master/fifa/fifa_countries_audience.csv"

fifa <- read.csv(data_url)
```

## Exploratory Data Analysis
Adding in a cleaner `continent` column based on a country's confederation. These loosely represent the geographic location of [FIFA confederations](https://en.wikipedia.org/wiki/List_of_football_federations)
```{r}
# Adding continent column for cleaner plots/readability
confederation <- c("CONCACAF", "UEFA", "CONMEBOL", "AFC", "CAF", "OFC")
continent <- c("North America", "Europe", "South America", "Asia", "Africa", "Oceania")
continent_map <- data.frame(confederation, continent)

fifa <- merge(fifa, continent_map)
```


We'll need to look at the `gdp_weighted_share` variable in our dataframe as that accounts for the population/GDP of a country. First, to compare the viewership of European countries vs that of non-European countries, we'll need to group our data into European vs non-European nations. We'll also wa
```{r}
fifa <- fifa %>%
              mutate(european = ifelse(fifa$confederation == "UEFA", TRUE, FALSE))
```

Let's take a look at some of the summary stats for our dataset via the `summary` method
```{r}
summary(fifa)
```

#### Data Visualization
Adding some plots to paint a better picture of our data. Let's start with the `gdp_weighted_share` variable
```{r}
# GDP weighted viewership share
ggplot(fifa, aes(x = gdp_weighted_share)) + geom_histogram()
```
```{r}
# population share
ggplot(fifa, aes(x = population_share)) + geom_histogram()
```

```{r}
# TV audience share
ggplot(fifa, aes(x = tv_audience_share)) + geom_histogram()
```



One thing that'd be interested to see is the world map with countries colored by their `tv_audience_share` values. We'll need to grab latitude and longitudinal data for each country to create a choropleth map. We can use the `maps` library and the `map_data` function to grab a listing of countries with their coordinates. From there we can create choropleths with colormaps representing the metrics we care about (`gdo_weighted share`, `tv_audience_share`, etc).
```{r world-map-tv-share}
world_map <- map_data("world") %>%
              rename("country" = "region")

# USA named differently in each dataset, replacing to standardize column values (same wqith UK)
world_map$country <- str_replace(world_map$country, "USA", "United States")
world_map$country <- str_replace(world_map$country, "UK", "United Kingdom")

# Joining our world_map geographic data to our World Cup viewership data by country
fifa_coords <- world_map %>% 
  left_join(fifa, by = c("country"))

head(fifa_coords)
```

First, let's take a look at the FIFA Confederation membership of each country
```{r, confederation-membership}
ggplot(fifa_coords, aes(long, lat, group = group)) +
  geom_polygon(aes( group=group, fill=confederation)) + 
  ggtitle("FIFA Confederation memberdship by country") +
  xlab("Longitude (deg)") + ylab("Latitude (deg)")
```


```{r}
# Plotting TV Audience share by country
ggplot(fifa_coords, aes(long, lat, group = group)) +
  geom_polygon(aes( group=group, fill=tv_audience_share)) + 
  ggtitle("TV Audience share (%) of world cup viewership by country") +
  xlab("Longitude (deg)") + ylab("Latitude (deg)")
```
```{r, gdp-weighted-share-choropleth}
ggplot(fifa_coords, aes(long, lat, group = group)) +
  geom_polygon(aes( group=group, fill=gdp_weighted_share)) + 
  ggtitle("GDP-weighted share (%) of world cup viewership by country") +
  xlab("Longitude (deg)") + ylab("Latitude (deg)")

```


### Inference
Let's run our t-test comparing the gdp-adjusted viewership for european vs non-european countries'. We'll be conducting this test with a significance level $\alpha = 0.05$. Our null and alternative hypotheses are listed below:

- $H_0$: The mean GDP-weighted viewership share of the world cup in Europe is *not higher*  than that of other confederations
- $H_a$: The mean GDP-weighter viewership share of the world cup is *higher* than that of other confederations

```{r}
# Filtering into our two groups: Europe vs not Europe.
europe <- fifa %>% filter(european== TRUE)
other_countries <- fifa %>% filter(european == FALSE)

# Running one-tailed t-test using R built-in
t.test(europe$gdp_weighted_share, other_countries$gdp_weighted_share, alternative="greater")
```

Since our p-value is less than $\alpha = 0.05$, we can reject the null hypothesis and claim that GDP-adjusted average world cup viewership in Europe is higher than that of non-European countries.

Let's also run an inference test on our `tv_audience_share` variable ($\alpha = 0.05$) to see if there's a significant difference between European and non-European countries. We'll start with our hypotheses:
- $H_0$: The mean tv audience viewership share of the world cup in Europe is *not higher*  than that of other confederations
- $H_a$: The mean tv audience viewership share of the world cup is *higher* than that of other confederations
```{r}
# Running one-tailed t-test using R built-in
t.test(europe$tv_audience_share, other_countries$tv_audience_share, alternative="greater")
```
Our p-value of $0.4279 > \alpha$ indicates that we cannot reject the null hypothesis in this instance. There is no statistically significant different in raw TV audience viewership share between European and non-European countries

#### ANOVA Between Confederations
Lastly, it'd be interesting to run an ANOVA analysis on `tv_audience_share` between Confederations (not just Europe vs Non-Europe) to see if there's any significant differences in World Cup viewership between confederations. 

- $H_0$: There is no difference in mean tv audience World Cup viewership share between confederations
- $H_a$: The mean tv audience viewership share of the world cup is different between confederations

We will conduct our ANOVA testing at a 5% significance level ($\apha = 0.05$)

```{r, anova-tv-audience}
confederation_test <- aov(tv_audience_share ~ confederation, data = fifa)

summary(confederation_test)
```

Since our p-value (0.0244) is less than $\alpha = 0.05$, we can reject the null hypothesis and assert that there *is* a statistically significant different in mean TV audience viewership between FIFA confederations. 


### References
- [@break-fifa-538]




