---
title: "Introduction to linear regression"
author: ""
output:
  pdf_document: default
  html_document:
    includes:
      in_header: header.html
    css: ./lab.css
    highlight: pygments
    theme: cerulean
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: console
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(eval = TRUE, message = FALSE, warning = FALSE)
```

The Human Freedom Index is a report that attempts to summarize the idea of "freedom" through a bunch of different variables for many countries around the globe. It serves as a rough objective measure for the relationships between the different types of freedom - whether it's political, religious, economical or personal freedom - and other social and economic circumstances. The Human Freedom Index is an annually co-published report by the Cato Institute, the Fraser Institute, and the Liberales Institut at the Friedrich Naumann Foundation for Freedom.

In this lab, you'll be analyzing data from Human Freedom Index reports from 2008-2016. Your aim will be to summarize a few of the relationships within the data both graphically and numerically in order to find which variables can help tell a story about freedom.

## Getting Started

### Load packages
            
In this lab, you will explore and visualize the data using the **tidyverse** suite of packages. The data can be found in the companion package for OpenIntro resources, **openintro**.

Let's load the packages.

```{r load-packages, message=FALSE}
library(tidyverse)
library(openintro)
data('hfi', package='openintro')
```


### The data

The data we're working with is in the openintro package and it's called `hfi`, short for Human Freedom Index.

1.  What are the dimensions of the dataset?

We can see the dataframe fields and dimensions using R's built-in `colnames` and `dim.data.frame` functions.
```{r}
colnames(hfi)

dim.data.frame(hfi)
```


2.  What type of plot would you use to display the relationship between the personal
    freedom score, `pf_score`, and one of the other numerical variables? Plot 
    this relationship using the variable `pf_expression_control` as the predictor.
    Does the relationship look linear? If you knew a country's `pf_expression_control`,
    or its score out of 10, with 0 being the most, of political pressures and 
    controls on media content, would you be comfortable using a linear model to
    predict the personal freedom score?
I would use a scatter plot to show a relationship between `pf_score` and another numerical variable. First, we'll need to load the `ggplot2` plotting library
```{r}
library(ggplot2)
```

```{r}
ggplot(hfi, aes(x=pf_expression_control, y=pf_score)) + geom_point()
```
Looking at our scatter plot, there does appear to be a rough linear relationship between these two variables. In other words, as `pf_expression_control` increases, `pf_score` also increases linearly. I would be comfortable using a linear model to predict a personal freedom scroe based on the political expression score.

If the relationship looks linear, we can quantify the strength of the relationship with the correlation coefficient.

```{r cor}
hfi %>%
  summarise(cor(pf_expression_control, pf_score, use = "complete.obs"))
```

Here, we set the `use` argument to "complete.obs" since there are some observations of NA.

## Sum of squared residuals

<div id="boxedtext">
In this section, you will use an interactive function to investigate what we mean by "sum of squared residuals". You will need to run this function in your console, not in your markdown document. Running the function also requires that the `hfi` dataset is loaded in your environment.
</div>

Think back to the way that we described the distribution of a single variable. Recall that we discussed characteristics such as center, spread, and shape. It's also useful to be able to describe the relationship of two numerical variables, such as `pf_expression_control` and `pf_score` above.

3.  Looking at your plot from the previous exercise, describe the relationship 
    between these two variables. Make sure to discuss the form, direction, and 
    strength of the relationship as well as any unusual observations.
    
There is a postiviely-increasing relationship between these two variables. That is, as `pf_expression_control` increases, the corresponding `pf_score` of a country also increases. While a linear model could be used here, there is a fair amount of spread in terms of `pf_score` for countries with a given `pf_expression_control` score. Let's take a look at the largest spread for countries in this `pf_expression_control` "band

```{r}
hfi %>%
    filter(pf_expression_control==5) %>%
    summarise(min_pf = min(pf_score), max_pf = max(pf_score))


```
We see a difference of over 5 points in `pf_score` between two countries with the same expression score.

In addition, there are several points towards the bottom left of this plot that do not follow the general trend of the plot. They have a lower `pf_score` than we would typically expect for their corresponding `pf_expression_control` score.

Just as you've used the mean and standard deviation to summarize a single variable, you can summarize the relationship between these two variables by finding the line that best follows their association. Use the following interactive function to select the line that you think does the best job of going through the cloud of points.

```{r plotss-expression-score, eval=FALSE}
# This will only work interactively (i.e. will not show in the knitted document)
hfi <- hfi %>% filter(complete.cases(pf_expression_control, pf_score))
DATA606::plot_ss(x = hfi$pf_expression_control, y = hfi$pf_score)
```

After running this command, you'll be prompted to click two points on the plot to define a line. Once you've done that, the line you specified will be shown in black and the residuals in blue. Note that there are 30 residuals, one for each of the 30 observations. Recall that the residuals are the difference between the observed values and the values predicted by the line:

\[
  e_i = y_i - \hat{y}_i
\]

The most common way to do linear regression is to select the line that minimizes the sum of squared residuals. To visualize the squared residuals, you can rerun the plot command and add the argument `showSquares = TRUE`.

```{r plotss-expression-score-squares, eval=FALSE}
DATA606::plot_ss(x = hfi$pf_expression_control, y = hfi$pf_score, showSquares = TRUE)
```

Note that the output from the `plot_ss` function provides you with the slope and intercept of your line as well as the sum of squares.

4.  Using `plot_ss`, choose a line that does a good job of minimizing the sum of
    squares. Run the function several times. What was the smallest sum of 
    squares that you got? How does it compare to your neighbors?
My smallest sum of squares was 971.021. It's in a similar range to my other trials as I attempted to select points that would form a line through the middle of the "cloud"

## The linear model

It is rather cumbersome to try to get the correct least squares line, i.e. the line that minimizes the sum of squared residuals, through trial and error. Instead, you can use the `lm` function in R to fit the linear model (a.k.a. regression line).

```{r m1}
m1 <- lm(pf_score ~ pf_expression_control, data = hfi)
```

The first argument in the function `lm` is a formula that takes the form `y ~ x`. Here it can be read that we want to make a linear model of `pf_score` as a function of `pf_expression_control`. The second argument specifies that R should look in the `hfi` data frame to find the two variables.

The output of `lm` is an object that contains all of the information we need about the linear model that was just fit. We can access this information using the summary function.

```{r summary-m1}
summary(m1)
```

Let's consider this output piece by piece. First, the formula used to describe the model is shown at the top. After the formula you find the five-number summary of the residuals. The "Coefficients" table shown next is key; its first column displays the linear model's y-intercept and the coefficient of `pf_expression_control`. With this table, we can write down the least squares regression line for the linear model:

\[
  \hat{y} = 4.61707 + 0.49143 \times pf\_expression\_control
\]

One last piece of information we will discuss from the summary output is the Multiple R-squared, or more simply, $R^2$. The $R^2$ value represents the proportion of variability in the response variable that is explained by the explanatory variable. For this model, 63.42% of the variability in runs is explained by at-bats.

5.  Fit a new model that uses `pf_expression_control` to predict `hf_score`, or
    the total human freedom score. Using the estimates from the R output, write 
    the equation of the regression line. What does the slope tell us in the 
    context of the relationship between human freedom and the amount of political
    pressure on media content?
    
First, let's generate our linear model:
```{r}
m2 <- lm(hf_score ~ pf_expression_control, data = hfi)
```

```{r}
summary(m2)
```
Using our ourput model from `m2`, we can write the equation of our linear model between `hf_score` and `pf_expressionc_control` as:

\[
  \hat{y} = 5.153687 + 0.349862 \times pf\_expression\_control
\]

The slope of our model (0.349862) tells us the increase we should expect in our `hf_score` given an increase in `pf_expression_control`. In other words, as political pressure exerted on a country's media is lessened, and the expression score increases, the human freedom score associated with that country will increase as well.

## Prediction and prediction errors

Let's create a scatterplot with the least squares line for `m1` laid on top.

```{r reg-with-line}
ggplot(data = hfi, aes(x = pf_expression_control, y = pf_score)) +
  geom_point() +
  stat_smooth(method = "lm", se = FALSE)
```

Here, we are literally adding a layer on top of our plot. `geom_smooth` creates the line by fitting a linear model. It can also show us the standard error `se` associated with our line, but we'll suppress that for now.

This line can be used to predict $y$ at any value of $x$. When predictions are made for values of $x$ that are beyond the range of the observed data, it is referred to as *extrapolation* and is not usually recommended. However, predictions made within the range of the data are more reliable. They're also used to compute the residuals.

6.  If someone saw the least squares regression line and not the actual data, how
    would they predict a country's personal freedom school for one with a 6.7 rating
    for `pf_expression_control`? Is this an overestimate or an underestimate, and
    by how much? In other words, what is the residual for this prediction?
    
They would input an expression score of 6.7 into the linear model equation above to determine the y-value (`pf_score`)

```{r}
y_pred <- 4.61707 + (0.49143 * 6.7)
y_pred
```

So we would expect a country with a `pf_expression_score` to have a `pf_score` of 7.909651. The residual for this prediction would be 0, as it lies along the model line and would have no distance (residual) between the prediction and model.


## Model diagnostics

To assess whether the linear model is reliable, we need to check for (1) linearity, (2) nearly normal residuals, and (3) constant variability.

**Linearity**: You already checked if the relationship between `pf_score` and `pf_expression_control` is linear using a scatterplot. We should also verify this condition with a plot of the residuals vs. fitted (predicted) values.

```{r residuals}
ggplot(data = m1, aes(x = .fitted, y = .resid)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  xlab("Fitted values") +
  ylab("Residuals")
```

Notice here that `m1` can also serve as a data set because stored within it are the fitted values ($\hat{y}$) and the residuals. Also note that we're getting fancy with the code here. After creating the scatterplot on the first layer (first line of code), we overlay a horizontal dashed line at $y = 0$ (to help us check whether residuals are distributed around 0), and we also reanme the axis labels to be more informative.

7.  Is there any apparent pattern in the residuals plot? What does this indicate
    about the linearity of the relationship between the two variables?  
    
There does not seem to be a trend/relationship in the residual plot (i.e., a slope of 0 or nearly 0 between our residuals and fitted values). This indicates that our predicted values are in line with the data. In addition, the residuals appear to be centered around 0, so our model is not "offset" in comparison to the data.

</br>

**Nearly normal residuals**: To check this condition, we can look at a histogram

```{r hist-res}
ggplot(data = m1, aes(x = .resid)) +
  geom_histogram(binwidth = 0.25) +
  xlab("Residuals")
```

or a normal probability plot of the residuals.

```{r qq-res}
ggplot(data = m1, aes(sample = .resid)) +
  stat_qq()
```

Note that the syntax for making a normal probability plot is a bit different than what you're used to seeing: we set `sample` equal to the residuals instead of `x`, and we set a statistical method `qq`, which stands for "quantile-quantile", another name commonly used for normal probability plots.

8.  Based on the histogram and the normal probability plot, does the nearly 
    normal residuals condition appear to be met?
The normal probability plot appears roughly linear (with some tailing off at the edges). In addition, the histogram appears nearly normal from an eye test (centered at 0, symmetrical), so we can check that condition off.

</br>

**Constant variability**:

9.  Based on the residuals vs. fitted plot, does the constant variability condition 
    appear to be met?
There does appear to be relatively constant variability in the fitted values vs residuals plot. There are fewer observations at the higher end of the plot so there does appear to be smaller variability there. Overall, however, it does appear to not change variability between these two parameters.
* * *

## More Practice
<div class="oyo">

-   Choose another freedom variable and a variable you think would strongly
    correlate with it. Produce a scatterplot of the two variables and fit a 
    linear model. At a glance, does there seem to be a linear relationship?
    
I'd like to take a look at the relationship between a country's hf_score and the freedom of movement score `pf_movement`. Let's first construct a linear model and plot. There does seem to be a linear relationship between the two variables.
```{r}
m3 = lm(hf_score ~ pf_movement, data = hfi)


ggplot(hfi, aes(x = pf_movement, y = hf_score)) +
      geom_point()  +
  stat_smooth(method = "lm", se = FALSE)

summary(m3)
```
Our linear model between these two variables has a coefficient of 0.279320 and intercept of 4.8082. 

-   How does this relationship compare to the relationship between 
    `pf_expression_control` and `pf_score`? Use the $R^2$ values from the two 
    model summaries to compare. Does your independent variable seem to predict
    your dependent one better? Why or why not?
    
Our relationship between freedom of movement scores and human freedom scores has an $R^2 = 0.5158$. This represents a weaker correlation between variables than the one from above. If we switch the independent and dependent variable we get the results below:

```{r}
m3_inverted = lm(pf_movement ~ hf_score, data = hfi)

summary(m3_inverted)
```
Which has the same correlation as the ones above, but different intercepts and residuals. This makes sense as the correlation coefficient only represents the strength of a relationship between two variables. In this case, our `pf_movement` variable is a better predictor of our `hf_score` variable than vice versa. 


-   What's one freedom relationship you were most surprised about and why? Display
    the model diagnostics for the regression model analyzing this relationship.
    
I'm interested in looking at the relationship between the number of violent conflicts, captured in the `pf_ss_disappearances_violent` field and a country's access to the internet (`pf_expression_internet`):
```{r}
m4 <- lm(pf_ss_disappearances_violent ~ pf_expression_internet, data = hfi)

summary(m4)
```
We have a correlation coefficient $R^2 = 0.01035$ which does not show a strong relationship. Let's create a scatter plot of these two variables to confirm.

```{r}
ggplot(hfi, aes(x = pf_expression_internet, y = pf_ss)) +
  geom_point() + 
  stat_smooth(method = "lm", se = FALSE)
  

```

* * *
